---
title: "Problem Set 1: Solutions"
author: "Your Name - Net ID - Section Number "
date: "Due Oct 6th, 2023"
output:
  pdf_document: default
---

This homework must be turned in on Brightspace by Oct 6th 2023. It must be your own work, and your own work only -- you must not copy anyone's work, or allow anyone to copy yours. This extends to writing code. \textbf{You may consult with others, but when you write up, you must do so alone.}

Your homework submission must be written and submitted using Rmarkdown. No handwritten solutions will be accepted. You should submit:

1. A compiled PDF file named yourNetID_solutions.pdf containing your solutions to the problems.

2. A .Rmd file containing the code and text used to produce your compiled pdf named yourNetID_solutions.Rmd.


Note that math can be typeset in Rmarkdown in the same way as Latex. Please make sure your answers are clearly structured in the Rmarkdown file:

1. Label each question part

2. Do not include written answers as code comments. Write out answers and explanations separately.

3. The code used to obtain the answer for each question part should accompany the written answer. Comment your code!

\newpage 

# Question 1. Definitions and Examples (20 points)
Answer the following questions. Be as specific and detailed as possible. Give examples.

1. What is the fundamental problem of causal inference? (5 points)

The fundamental problem of causal inference is that it is impossible to observe the individual treatment effect and therefore it is challenging to determine the causal effect of a treatment. This issue mainly stems from the fact that it's impossible to observe both potential outcomes of treatment and control; if the unit receives the treatment it is impossible to observe the individual effect of the counterfactual (receiving no treatment) and vice versa.

2. Why are experiments important? (5 points)

Experiments, especially randomized controlled trials (RCTs) are the fundamental "solution" to the fundamental problem of causal inference. Experiments are studies where the researcher controls the probability of being assigned the treatment. One can engineer the experiment to meet the requirements for the difference-in-means to be equal to average treatment effect (ATE). The three requirements are SUTVA, positivity, and ignorability. When these requirements are satisfied, 

3. What does ignorability mean? (5 points)

Ignorability, or unconfoundedness refers to the treatment assignment being independent of the potential outcomes given some observed covariates. In other words, whether or not a case is assigned to treatment will not affect the case's potential outcome. Here is an example of an ignorability violation: suppose you were carrying out a study 

4. What is SUTVA? (5 points)

SUTVA stands for Stable Unit Treatment Value Assumption. It is one of the requirements for the difference-in-means to be equal to the ATE. SUTVA calls for no spillover (or interference) and single value of treatment. No spillover means one unit's treatment status will not affect the potential outcome of other units. An example of a spillover violation is this: let's say you had a study on the efficacy of having a scarecrow on pest-related crop damage. You put a scarecrow in one field as the treatment, and leave a nearby nearly identical field untouched as control. However, since the scarecrow does manage to scare away some pestilent animals, these affected animals instead migrate to the scarecrow-free field increasing the pest damage in the control field. This is an example of spillover since the treatment affected the outcome of the control group. Single value of treatment refers to a single version of treatment for all treatments of the same level; in other words the process by which the treatment is given does not affect the treatment potential outcome. Here's an example of a single treatment value violation: imagine you conducted a study on the effect of psychological therapy on subjects' sense of fulfillment (only on mentally healthy people for obvious ethical reasons). You randomly assign subjects to treatment (therapy sessions) and control (no therapy), however you fail to standardize the actual therapy experience. Instead, subjects assigned to different therapists effectively receive different therapy treatments. As a result, those who were lucky enough to be assigned to more skilled or experienced therapists had "better" outcomes (higher sense of fulfillment) than those assigned to inexperienced or unenthusiastic therapists. This is a violation of single treatment value as differences in how exactly the treatment was received affected the individual treatment outcomes.

\newpage



# Question 2. Bed Nets and Malaria (20 points)

Article: Free Distribution or Cost-Sharing? Evidence from a Randomized Malaria Prevention Experiment
by Jessica Cohen and Pascaline Dupas

Some economists have argued that ``cost-sharing" makes it more likely that a product will be used (versus giving it away for free). Cohen and Dupas partnered with 20 Kenyan prenatal clinics to distribute subsidized anti-malarial bed nets. For each clinic, they varied the extent of the subsidy: either full (free bed-nets, $D_i = 1$) or partial (90\% cheaper bed-nets, $D_i = 0$). They measure (among other things) whether women who received bed nets used them ($Y_i$).


1. What is $\mathbb{E}[Y_i | D_i = 0]$? (5 points)



2. What is $\mathbb{E}[Y_i(1)]$? (5 points)



3. What is $\mathbb{E}[Y_i(1) | D_i = 0]$? (5 points)



4. Cohen and Dupas randomized treatment at the level of the clinic, but the outcomes of interest are at the individual level. Is there a violation of consistency/SUTVA? Why or why not? Argue your case. (5 points)




\newpage

# Question 3. Application (Coding) (30 points)
The STAR (Student-Teacher Achievement Ratio) Project is a four year
*longitudinal study* examining the effect of class size in early
grade levels on educational performance and personal
development.

This exercise is in part based on\footnote{ I have provided you with a 
sample of their larger dataset. Empirical conclusion drawn from this 
sample may differ from their article.}:

 Mosteller, Frederick. 1997. “[The Tennessee Study of Class Size in the 
 Early School Grades.](http://dx.doi.org/10.2307/3824562)” *Bulletin of 
 the American Academy of Arts and Sciences* 50(7): 14-25.
  
A longitudinal study is one in which the same
participants are followed over time.  This particular study lasted
from 1985 to 1989 involved 11,601 students. During the four years of
the study, students were randomly assigned to small classes,
regular-sized classes, or regular-sized classes with an aid.  In all,
the experiment cost around $12 million. Even though the program
stopped in 1989 after the first kindergarten class in the program
finished third grade, collection of various measurements (e.g.,
performance on tests in eighth grade, overall high school GPA)
continued through the end of participants' high school attendance.

We will analyze just a portion of this data to investigate whether the
small class sizes improved performance or not. The data file name is
`STAR.csv`, which is a CSV data file.  The names and
descriptions of variables in this data set are:


 Name                 Description
 -------------------- ----------------------------------------------------------
 `race`               Student's race (White = 1, Black = 2, Asian = 3, Hispanic = 4,  Native American = 5, Others = 6)
 `classtype`          Type of kindergarten class (small = 1, regular = 2, regular with aid = 3)
 `g4math`             Total scaled score for math portion of fourth grade standardized test 
 `g4reading`          Total scaled score for reading portion of fourth grade standardized test 
 `yearssmall`         Number of years in small classes 
 `hsgrad`             High school graduation (did graduate = 1, did not graduate = 0) 
 
Note that there are a fair amount of missing
values in this data set.  For example, missing values arise because
some students left a STAR school before third grade or did not enter a
STAR school until first grade.

1. Create a new factor variable called `kinder` in the data
  frame.  This variable should recode `classtype` by changing
  integer values to their corresponding informative labels (e.g.,
  change 1 to `small` etc.).  Similarly, recode the
  `race` variable into a factor variable with four levels
  (`white`, `black`, `hispanic`, `others`) by
  combining Asians and Native Americans as the `others`
  category.  For the `race` variable, overwrite the original
  variable in the data frame rather than creating a new one.  Recall
  that `na.rm = TRUE` can be added to functions in order to
  remove missing data. (5 points)

```{r}
library("tidyverse")
STAR <- read.csv("STAR2.csv")

STAR <- STAR %>% 
  mutate(kinder = factor(case_when(classtype == 1 ~ "small",
                                   classtype == 2 ~ "regular",
                                   .default = "regular with aid")),
         race = factor(case_when(race == 1 ~ 'white',
                                 race == 2 ~ 'black',
                                 race == 4 ~ 'hispanic',
                                 .default = 'others'))) 

```

2. How does performance on fourth grade reading and math tests for
  those students assigned to a small class in kindergarten compare
  with those assigned to a regular-sized class?  Do students in the
  smaller classes perform better?  Use means to make this comparison
  while removing missing values.  Give a brief substantive
  interpretation of the results.  To understand the size of the
  estimated effects, compare them with the standard deviation of the
  test scores. (10 points)
 
```{r}

STAR %>%
  group_by(kinder) %>%
  summarise(`reading test mean` = mean(g4reading, na.rm = T),
            `math test mean` = mean(g4math, na.rm = T),
            `reading test sd` = sd(g4reading, na.rm = T),
            `math test sd` = sd(g4math, na.rm = T))
 
```






3. Instead of comparing just average scores of reading and math
  tests between those students assigned to small classes and those
  assigned to regular-sized classes, look at the entire range of
  possible scores.  To do so, compare a high score, defined as the
  66th percentile, and a low score (the 33rd percentile) for small
  classes with the corresponding score for regular classes.  These are
  examples of *quantile treatment effects*.  Does this analysis
  add anything to the analysis based on mean in the previous question? 
  (Hint: You will use the quantile() function in r.) (5 points)

```{r}

STAR %>%
  group_by(kinder) %>%
  summarise(`reading test 33rd pct` = quantile(g4reading, 0.33, na.rm = T),
            `reading test 66th pct` = quantile(g4reading, 0.66, na.rm = T),
            `math test 33rd pct` = quantile(g4math, 0.33, na.rm = T),
            `math test 66th pct` = quantile(g4reading, 0.66, na.rm = T))

```

4. We examine whether the STAR program reduced the achievement gaps
  across different racial groups.  Begin by comparing the average
  reading and math test scores between white and minority students
  (i.e., Blacks and Hispanics) among those students who were assigned
  to regular classes with no aid.  Conduct the same comparison among
  those students who were assigned to small classes.  Give a brief
  substantive interpretation of the results of your analysis. (5 points)
  
```{r}

STAR %>%
  filter(kinder != 'regular with aid') %>%
  group_by(kinder, race) %>%
  summarise(`reading test mean` = mean(g4reading, na.rm = T),
            `math test mean` = mean(g4math, na.rm = T))

```

5. We consider the long term effects of kindergarten class size.
  Compare high school graduation rates across students assigned to
  different class types.  Also, examine whether graduation rates
  differ by the number of years spent in small classes.  Finally, as
  done in the previous question, investigate whether the STAR program
  has reduced the racial gap between white and minority students'
  graduation rates.  Briefly discuss the results. (5 points)

```{r}

STAR %>%
  group_by(kinder) %>%
  summarise(`graduation rate` = mean(hsgrad, na.rm = T))

```
```{r}

STAR %>%
  group_by(yearssmall) %>%
  summarise(`graduation rate` = mean(hsgrad, na.rm = T))

```

\newpage
## Question 4. Design Your Experiment (30 points)

Design your own experiment from start to finish. Choose an *interesting* question. Explain why observational data may give you the wrong answer. Detail the potential outcomes and a well-defined treatment. Explain the type of experiment (completely random, cluster-design, block/stratified). Will your design ensure a causal treatment effect? (Remember: Be as specific as possible and give examples.)
